\documentclass[pdftex,aspectratio=169,14pt]{beamer}
\input{lecture-setup}
\lstMakeShortInline[language=HACS,columns=fullflexible]"

%% Style.
\usepackage[T1]{fontenc}
\usepackage{charter}
\renewcommand{\ttdefault}{txtt}

\useinnertheme{circles}
\useoutertheme{default}
\usefonttheme{serif}
%\SelectTips{lu}{10 scaled 2074}

%% Topmatter
\title{ \hax }
\author{ Kristoffer H. Rose \\
 Two Sigma Investments, LLC \& New York University
}
\date{September 8, 2015 \\
  CHUM 4 (IBM Watson, Yorktown Heights)
}

\begin{document}
\maketitle
%\begin{frame}{Outline}
%  \tableofcontents
%\end{frame}

\section{Syntax}
%\begin{frame}{Outline}
%  \tableofcontents[current]
%\end{frame}

\begin{frame}{\hax syntax}\small
  \vspace*{-1ex}
  \begin{align}
    \tag{\hax{}Script}
    H &::= \ov{D} 
    \\
    \tag{Declaration}
    D &::= S~\kw{data}~d\,\kw(\,\ov{F}\,\kw)\,\kw;
    \bigm| S~\kw{scheme}~f\,\kw(\,\ov{F}\,\kw)\,\kw;
    \bigm| S~\kw{variable}\,\kw;
    \bigm| S~\kw{rule}~M~\kw{$→$}~M\,\kw;
    \\
    \tag{Form}
    F &::= \kw[\,\ov{S}\,\kw]\,S
    \bigm| \kwm\{\, S:S \,\kwm\}
    \\
    \tag{Sort}
    S &::= s\,\kwm{⟨}\,\ov{S}\,\kwm{⟩}
    \bigm| α
    \\[\jot]
    \tag{Meta-Term}
    M &::= c\,\kw(\,\ov{P}\,\kw)
    \bigm| v
    \bigm| m\,\kw(\,\ov{M}\,\kw)
    \\
    \tag{Piece}
    P &::= \kw[\,\ov{v}\,\kw]\,M
    \bigm| \kwm\{\, \ov{A} \,\kwm\}
    \\
    \tag{Association}
    A &::= M\,\kw:\,M
    \bigm| \kw{$¬$}\,M
    \bigm| \kw\,m\,\kw(\,\ov{v}\,\kw)
  \end{align}
\end{frame}

\begin{frame}[fragile]{\hax type system}\small
  \begin{enumerate}

  \item All rules collected under top sort.

  \item Each \structure{scheme} function can have unique attributes.

  \item All \structure{data} constructors in sort must have \emph{same} attributes.

  \item \emph{All} terms must have all attributes.

  \item Map attributes are \emph{only} from variables of key sort, which must include \kw{variable}.

  \item Meta-variables can \emph{only} have data attribute constraints for their sort.
    \begin{itemize}
    \item in patterns this implements ``Data'' option…
    \end{itemize}

  \item "{¬…}" only in patterns, "{}" only in contractions.

  \item Substitution is \emph{only} of bound variables where binder has sort, which must include
    \kw{variable}.

  \end{enumerate}
\end{frame}

\begin{frame}{\HAX to \hax}
  \begin{enumerate}

  \item Resolve all syntax.

  \item \structure{Regroup} all sorted declarations into top sort.

  \item Complete each term instance with \structure{all possible attributes}.
    \begin{itemize}
    \item eliminates all ``catch-all'' attribute meta-variables.
    \end{itemize}

  \item Regular attributes converted to map attributes from dummy variable sort.

  \item Erase synthesized and inherited ("↑" and "↓") markers.

  \item \alert{Something about rule strategy.}

  \end{enumerate}
\end{frame}

\begin{frame}{\CRSX to \hax}
  \begin{enumerate}

  \item Resolve all syntax.

  \item All environments are resolved into---
    \begin{itemize}
    \item Top layer with \structure{constant→map}.
    \item Second layer (map) is  \structure{variable→value} (dummy as needed).
    \end{itemize}

  \item Data declaration becomes dummy data attribute.

  \end{enumerate}
  \alert{Incomplete!}
\end{frame}

\end{document}


%------------------------------------------------------------------------
% Tell Emacs that this is a LaTeX document and how it is formatted:
% Local Variables:
% mode:latex
% fill-column:100
% TeX-master: t
% End:
